\documentclass[sigconf,nonacm]{acmart}

%% Axiomander -- Iris experience report
%% ACM acmart class, sigconf (two-column conference) format.
%% nonacm: this is a self-archived draft, not a submitted ACM proceedings copy.

\usepackage{booktabs}
\usepackage{mathtools}
\usepackage{mathpartir}
\usepackage{listings}
\usepackage{xcolor}

%% Separation-logic and Iris notation
\newcommand{\sepc}{\mathbin{\ast}}
\newcommand{\pto}{\mapsto}
\newcommand{\wand}{\mathrel{-\!\ast}}
\newcommand{\pure}[1]{\ulcorner #1 \urcorner}
\newcommand{\wpre}{\mathrm{wp}}
\newcommand{\lget}{\mathrm{lget}}
\newcommand{\lupd}{\mathrm{lupd}}
\newcommand{\clobber}{\mathrm{clobber}}
\newcommand{\hoare}[3]{\{#1\}\ #2\ \{#3\}}
\newcommand{\code}[1]{\texttt{#1}}
\let\Bbbk\relax
\usepackage{amssymb}

%% SnakeletLang / SnakeletWp notation
\newcommand{\later}{{\triangleright}}
\newcommand{\wpe}[2]{\mathrm{WPE}\ #1\ \{\!\!\{\, #2 \,\}\!\!\}}
\newcommand{\rval}[1]{\mathsf{val}\,#1}
\newcommand{\rexn}[2]{\mathsf{exn}\,#1\,#2}
\newcommand{\sstep}{\leadsto}
\newcommand{\subs}[3]{#1[#2{:=}#3]}
\newcommand{\entries}{\mathrm{entries}}
\newcommand{\binopd}[2]{\mathrm{binop}(\mathit{op},#1,#2)}

%% Lightweight code listing style for Python / Coq snippets
\lstdefinestyle{snippet}{
  basicstyle=\ttfamily\footnotesize,
  breaklines=true,
  columns=fullflexible,
  keepspaces=true,
  showstringspaces=false,
  frame=single,
  framesep=4pt,
  rulecolor=\color{gray!40},
  xleftmargin=2pt,
  xrightmargin=2pt,
}
\lstset{style=snippet}

\begin{document}

\title{From Flat Stores to Separation Logic:
Lessons Building a Python Verifier}
\subtitle{Experience report: migrating an LLM-assisted verifier to Iris}

\author{Gavin Mendel-Gleason}
\affiliation{%
  \institution{Scidonia}
  \country{}
}
\email{}

\renewcommand{\shortauthors}{Mendel-Gleason}

\begin{abstract}
Axiomander is a verification pipeline for ordinary, decorator-free Python: a
human writes a strong specification, an LLM proposes the implementation and the
helper contracts needed to prove it, and the verifier returns a deterministic
verdict with a structured residual on failure. This experience report is about
the proof backend underneath that loop. We first built it the obvious way --
lowering Python to a small imperative language over a flat store -- and recovered
the locality that modular verification needs by \emph{reconstructing the frame
rule by hand}: an explicit \code{clobber} operator and generated per-callee,
per-variable frame lemmas. It worked for tiny inputs and then stopped scaling:
weakest-precondition terms blew up, the compiled-Ltac automation grew brittle,
and whole classes of programs -- owned mutable fields, aliasing, anything heading
toward concurrency -- could not be expressed at all. We then rebuilt the backend
on Iris, where a callee's footprint is a separating resource and framing is the
separating conjunction. Our thesis, which the migration bore out, is that the
abstractions an LLM-driven verifier needs -- modular verification, proof reuse,
library stubbing, and incremental checking -- are all instances of one mechanism:
Iris framing behind opaque specification boundaries. We report what changed
quantitatively (framing obligations per call fell to zero; automatically
generated, machine-checked proofs replaced hand-decomposed ones) and
qualitatively (resource-bearing code became expressible at all), and distil the
lessons for others building automated, composition-heavy verifiers.
\end{abstract}

\keywords{separation logic, Iris, weakest preconditions, specification
composition, stubbing, frame rule, LLM-assisted proof, Python, experience report}

\maketitle

\section{Motivation: vericoding needs composition}
\label{sec:motivation}

Axiomander's development model is a loop. The human writes a \emph{strong
specification} -- a precondition/postcondition pair -- as the primary artifact.
An LLM proposes an implementation. The verifier reduces ``does the code meet the
spec?'' to a proof obligation and returns either a certificate or a
\emph{structured residual} (the remaining goal with its hypotheses, plus a
counterexample where one exists). The LLM repairs against the residual, not
against the original natural-language prompt. Crucially, the LLM has freedom in
\emph{sub-specification}: to prove an entry point it may introduce helper
functions, each with its own contract, which are checked the same way. The LLM is
the reason we need what follows, but it is not the subject of this report: the
subject is the proof backend that has to make the loop tractable.

For this loop to be usable on real codebases, three forms of composition must
hold simultaneously:

\begin{enumerate}
  \item \textbf{Modular verification.} A caller must be provable from a callee's
  \emph{contract} alone. If proving \code{f} required unfolding the bodies of
  everything \code{f} transitively calls, neither proofs nor LLM prompts would
  scale.
  \item \textbf{Proof reuse under iteration.} Editing a function body must not
  invalidate its callers' proofs as long as the function's \emph{contract} is
  unchanged. The verifier should behave like an incremental build system:
  \emph{body changes invalidate local proofs; contract changes invalidate
  callers.}
  \item \textbf{Stubbing.} Libraries we cannot or will not verify must be
  replaceable by their declared behaviour. A caller proves against the stub's
  contract, which becomes an assumption; the trust boundary is the stub, written
  and reviewed once.
\end{enumerate}

All three are, at bottom, the \textbf{frame property}: reasoning about one
component must be insensitive to the parts of the state it does not touch. The
thesis of this report is that these are not three problems but one, that an
off-the-shelf separation logic supplies it directly, and that the place we spent
our effort -- and hit our limits -- was trying to reconstruct it over a flat
store. The rest of the paper is that story: how we built framing the hard way
(Section~\ref{sec:imp}), why we abandoned it for Iris (Section~\ref{sec:iris}),
what measurably and expressively changed (Section~\ref{sec:eval}), and what we
would tell the next team (Section~\ref{sec:lessons}).

\section{First architecture: framing by hand over a flat store}
\label{sec:imp}

Our original backend lowers Python to a small imperative language IMP, with a
state that is a flat finite map from variable names to values
(\code{VZ\,|\,VBool\,|\,VUnit}, plus structural \code{VList\,|\,VTuple\,|\,VDict},
with mutation modelled on a parallel heap representation). The WP calculus is
standard and proven sound in Rocq:
\begin{align*}
\wpre(\mathsf{skip}, Q) &= Q, \\
\wpre(x := e, Q) &= Q[x \mapsto e], \\
\wpre(c_1; c_2, Q) &= \wpre(c_1, \wpre(c_2, Q)), \\
\wpre(\mathsf{if}\ e\ c_1\ c_2, Q) &= (e \Rightarrow \wpre(c_1,Q)) \\
  &\quad {}\wedge (\neg e \Rightarrow \wpre(c_2,Q)).
\end{align*}

Calls are the interesting case. A call site \code{t := g(a)} is verified against
\code{g}'s contract using a \code{CCall} rule. To recover locality -- to let a
caller conclude that variables it cares about survive the call -- we model the
callee's effect with an explicit \code{clobber} operator that havocs the callee's
declared write set, and we must discharge, for \emph{every} variable $v$, an
obligation of the form
\begin{multline*}
\forall v.\; v \notin (\textit{target} :: \textit{writes}) \Rightarrow \\
\lget\,s\,v = \lget\,(\clobber\,(\lupd\,s\,\textit{target}\,r)\,\textit{writes})\,v.
\end{multline*}

This is the frame rule, but encoded extensionally over a flat store. Three
problems follow, all of which we hit in practice. Two are about cost:

\begin{itemize}
  \item \textbf{WP-term blowup.} Each successive call nests the full WP expansion
  inside the previous call's postcondition. A function with a handful of
  sequential calls produces terms too large for the kernel to handle comfortably,
  and the monolithic proof tactic stops closing them.
  \item \textbf{Brittle automation.} The single universally-quantified frame
  subgoal does not match cleanly in compiled Ltac, partly because coercions
  (e.g.\ an \code{ls} wrapper) normalise away in the \code{.vo} and defeat the
  source-level pattern.
\end{itemize}

\noindent The third is about \emph{expressivity}, and it is the one we
underestimated: a flat name-to-value store cannot model two object references
that alias the same cell, cannot transfer ownership of a mutable field across a
call, and has no path to concurrency. Whole programs -- not just whole proofs --
fall outside it.

Our mitigation for the cost problems was to generate, at the Python IR level,
\textbf{one frame lemma per (callee, preserved-variable) pair}:

\begin{lstlisting}
(* g writes ["result"]; the caller's "a" is
   preserved across t := g(a) *)
Lemma g_frame_a : forall (s : state) (r : Z),
  ~ In "a" ("t" :: "result" :: nil) ->
  lget s "a" =
    lget (clobber (lupd s "t" (VZ r))
                  ("result" :: nil)) "a".
Proof. apply wp_ccall_frame. Qed.
\end{lstlisting}

The caller's proof then replaces the monolithic $\forall v$ subgoal with a
sequence of \code{apply g\_frame\_a. apply g\_frame\_b. ...}, each trivial and
each failing independently with its own residual. This restores tractability and
keeps the per-obligation, fail-locally discipline we want. But it is, frankly, a
workaround: we are \emph{reconstructing} the frame rule, one variable at a time,
in a logic that does not have it -- and reconstruction does nothing for the
expressivity gap. The lesson is the familiar one: \textbf{if framing is
load-bearing, the assertion logic should be substructural.}

\section{Migration to Iris}
\label{sec:iris}

We rebuilt the backend on Iris. \textbf{Snakelet} is the intermediate language
into which Axiomander lowers Python; \textbf{SnakeletWp} is its
weakest-precondition calculus, defined inside the Iris program logic. The state
is an Iris heap: a mutable Python object field is a points-to assertion
$l \pto v$, ownership is a separating resource rather than an entry in a flat
map, and assertions are \code{iProp}. Snakelet targets Python \emph{in the
large} -- ordinary control flow, exceptions, calls, and the built-in containers
lower directly, and only genuinely pathological constructs (runtime
metaprogramming, reflection) are gated with a diagnostic. Its full term language,
small-step semantics, and the machine-checked WP rule set are in
Appendix~\ref{app:snakelet}; this section is about what the move bought us.

\subsection{Contracts compile to \code{iProp}, composed with the separating
conjunction}

Contracts remain ordinary Python -- plain \code{assert} statements (leading $=$
precondition, trailing-before-return $=$ postcondition, in-loop $=$ invariant) or
a verifier-only \code{axiomander:} docstring. A shared \code{ContractLinter}
parses them into a contract IR; a backend-specific pass (\code{contract\_ir\_iris})
compiles that IR to \code{iProp}. Pure facts are injected with the persistent
embedding $\pure{P}$; resource facts compile to points-to assertions.

\begin{table}[t]
\caption{Contract fragments compiled to \code{iProp}.}
\label{tab:compile}
\small
\begin{tabular}{@{}ll@{}}
\toprule
Python contract fragment & \code{iProp} \\
\midrule
\code{assert x >= 1} & $\pure{x \ge 1}$ \\
\code{assert implies(A, B)} & $\pure{A \to B}$ \\
\code{assert all(x != i for i in range(0,10))} &
  $\pure{\forall i,\ 0 \le i < 10 \to x \ne i}$ \\
\code{owns(box)} & $\textit{box.value} \pto v$ (a resource) \\
\bottomrule
\end{tabular}
\end{table}

The decisive design choice is that \textbf{the default composition of contract
clauses is the separating conjunction $\sepc$, not $\wedge$.} When a precondition
or postcondition is built from several \code{assert}s, they are joined with
$\sepc$. For the pure-integer fragment this costs nothing, since
$\pure{P} \sepc \pure{Q} \dashv\vdash \pure{P \wedge Q}$; but it means the
\emph{same} composition rule already accounts for disjoint resources, and it is
the connective that lets us scale to invariants and atomic updates without
changing the front end.

A typical mixed pure/resource contract, written by the user with no Iris
notation:

\begin{lstlisting}[language=Python]
def bump(box: Box) -> int:
    """
    axiomander:
        requires:
            owns(box)
            box.value >= 0
        modifies:
            box.value
        ensures:
            owns(box)
            box.value == old(box.value) + 1
            result == box.value
    """
    box.value = box.value + 1
    return box.value
\end{lstlisting}

compiles to the Hoare triple
\begin{multline*}
\{\,\textit{box.value} \pto n \sepc \pure{n \ge 0}\,\}\quad
\textit{bump(box)} \\
\{\,r.\ \textit{box.value} \pto n{+}1 \sepc \pure{r = n{+}1}\,\}.
\end{multline*}

\code{owns(box)} plus \code{modifies: box.value} \emph{is} the resource
footprint; the user never writes $\sepc$, $\pto$, masks, or proof-mode tactics.
Note that this contract has no flat-store rendering at all: \code{owns} denotes
ownership of a mutable cell, which the name-to-value store of Section~\ref{sec:imp}
cannot express. Resource awareness is the first thing the migration \emph{adds},
before any question of proof size.

\subsection{The frame rule is free, and so is composition}

Because assertions are \code{iProp}, the structural frame rule
\[
\inferrule{\hoare{P}{c}{Q}}{\hoare{P \sepc R}{c}{Q \sepc R}}
\]
holds without a side condition: any resource $R$ describing a disjoint region of
the heap is carried across $c$ by the separating conjunction, automatically.
Compared with Section~\ref{sec:imp}, the entire \code{clobber}/$\forall v \notin
\textit{writes}$ apparatus, and the per-callee per-variable frame-lemma generator
that papered over it, simply disappear. Locality stops being something we encode
and becomes the shape of the logic.

\subsection{Opaque calls: one boundary, three jobs}
\label{sec:opaque}

Calls are dispatched through a \textbf{contract table} (an Iris \code{FunCtx})
that maps each callee to its specification. SnakeletWp provides two call rules: a
\emph{transparent} call unfolds a first-party body, and an \emph{opaque} call
consumes the callee's precondition (giving up the relevant resources), receives
its postcondition, and frames the rest. Schematically, for a callee with spec
$\hoare{P_g}{g}{Q_g}$:
\[
\inferrule{P_g \sepc R \\\\
  \forall r.\ (Q_g \sepc R) \vdash \wpre\ K\ \{\Phi\}}
{\wpre\ (\mathsf{call}\ g;\ K)\ \{\Phi\}}
\]

The single most useful thing we learned is that this one rule does three jobs at
once. It is \textbf{stubbing}: a library function given a contract -- in a
\code{.pyi} stub or an \code{axiomander:} docstring -- enters the \code{FunCtx}
and callers prove against it, body never required. It is \textbf{modular
verification}: a first-party callee can be re-declared opaque once its contract
is stable. And it is the \textbf{incremental-reuse boundary}: callers depend only
on the contract, so a body edit cannot disturb them, and the \code{FunCtx}
entry's hash is exactly the key our cache uses to decide what to re-prove.
Modularity, stubbing, and incrementality are not three features we built; they
are three readings of the opaque-call rule.

\subsection{Separation logic as an implementation detail}

The migration is deliberately confined to the proof backend. The
\code{ContractLinter}, the SMT export (\code{smt\_export}, \code{theory\_smt}),
the LLM oracle, and the function-level incremental cache are all
backend-agnostic; the contract IR diverges only at the final rendering step
(flat-store \code{to\_coq} vs.\ \code{contract\_ir\_iris.iris\_prop}). From the
user's side nothing changes: contracts are still decorator-free Python, and
\code{owns(...)} is the only new, entirely optional, vocabulary. The separating
conjunction, the points-to, the magic wand, and the \code{FunCtx} never surface.
Iris is the substrate, not the interface -- separation logic as an
\emph{implementation detail}.

\subsection{Staged, syntax-directed proofs}
\label{sec:staged}

SnakeletWp ships a small instruction set of \textbf{stage tactics} --
\code{pure\_step}, \code{case\_bool}, \code{call\_opaque},
\code{call\_transparent}, \code{finish\_pure}, and a precondition-discharge stage
\code{call\_opaque\_pre} -- and the proof generator (\code{iris\_proof\_gen})
emits exactly one stage per IR node, chosen by the node's syntax and the callee's
table entry. The generator performs no symbolic execution: each stage tactic
extracts what it needs from the Iris goal at proof time, so a three-call chain
becomes a flat, machine-checked sequence such as
\code{[call\_opaque, pure\_step, call\_transparent, pure\_step, pure\_step,
call\_opaque, finish\_pure]}. Crucially for the LLM loop, \textbf{stages fail
independently}: when \code{call\_opaque\_pre} cannot discharge a precondition the
residual is exported to SMT (an UNSAT result returns as \code{Axiom smt\_ax\_N},
the stage regenerated as \code{call\_opaque\_pre (exact smt\_ax\_N)}); a pure
postcondition \code{finish\_pure} cannot close is likewise an SMT candidate; and
beyond SMT the named stage plus its goal state -- not a stderr regex -- is what
the LLM oracle receives. A failed attempt yields a reusable artifact rather than
a dead end.

\section{Evaluation}
\label{sec:eval}

We compare the two backends on small functions drawn from the repository. Two
effects matter, and the second is the one we care about more: how much framing
the proof must do by hand, and what can be expressed at all.

\paragraph{Framing cost.} The flat-store backend pays for locality per call: at
each call site it must establish that every live variable outside the callee's
write set is preserved. For the two-call function \code{frame\_two\_calls} (two
increments and a sum) this means two explicit \code{wp\_ccall\_frame}
applications and fourteen preservation conjuncts threaded through intermediate
postconditions ($Q_1$: 6, $Q_2$: 8); the monolithic tactic does not close it, so
the proof is a hand-written staged decomposition. In Iris the frame rule carries
untouched resources implicitly, so the corresponding obligation count is
\emph{zero} -- independent of how many variables are live or how many calls
precede. Table~\ref{tab:eval} reports this against generator-emitted,
\code{coqc}-checked SnakeletWp proofs.

\begin{table}[t]
\caption{Framing cost and proof size on functions from the repository.
\emph{framing} counts explicit per-call frame steps plus preservation conjuncts
threaded through intermediate postconditions (``$2{+}14$'' $=$ 2 frame
applications $+$ 14 conjuncts); \emph{proof} is non-blank \code{Proof}--\code{Qed}
lines. Flat-store proofs are hand-decomposed (the monolithic tactic does not
scale past one call); Iris proofs are generated and \code{coqc}-checked.
``n/a'' $=$ not expressible over a flat store; ``---'' $=$ no scalable proof.}
\label{tab:eval}
\small
\begin{tabular}{@{}lcccc@{}}
\toprule
 & \multicolumn{2}{c}{flat store} & \multicolumn{2}{c}{Iris} \\
\cmidrule(lr){2-3}\cmidrule(lr){4-5}
function (calls) & framing & proof & framing & proof \\
\midrule
2-call chain        & $2{+}14$ & 29  & 0 & 11 \\
3-call chain        & ---      & --- & 0 & 14 \\
owned-field \code{bump}   & n/a & n/a & 0 & 15 \\
mutate-then-raise   & n/a      & n/a & 0 & \phantom{0}9 \\
\bottomrule
\end{tabular}
\end{table}

The trend, not the absolute numbers, is the point: flat-store framing grows with
(calls $\times$ live variables) and is discharged by hand, while the Iris column
is a flat zero and the proofs are emitted automatically. The three-call Iris
proof (14 lines) is \emph{shorter} than the two-call flat-store proof (29 lines)
despite doing more.

\paragraph{Expressivity, not just efficiency.} The bottom two rows are the more
important result. \code{bump} mutates an owned field; \code{mutate-then-raise}
mutates a cell and then raises, with the exceptional postcondition describing the
heap as it stood at the raise. Neither has a flat-store analogue at all -- the
name-to-value store cannot express ownership of a mutable cell, let alone
ownership surviving an exception. The migration did not merely make these proofs
shorter; it made the corresponding \emph{programs verifiable for the first time}.
The same observation extends to aliasing and to the concurrency primitives the
calculus already carries: separation logic widened what Axiomander can talk
about, and proof size is a secondary benefit.

\section{Discussion and related work}
\label{sec:discussion}

Axiomander's separation-logic interpretation of object fields and its opaque
call rule are textbook Iris; the novelty is not in the logic but in the
\emph{system context}, and the contribution of this report is the experience of
the migration rather than a new metatheory. To our knowledge the combination is
unusual: the assertion logic is never exposed to the user, the proof obligations
are generated and largely discharged by an LLM operating on structured residual
goals with Iris/Rocq as the trust kernel, and a single opaque-call rule serves as
modular verification, incremental-reuse boundary, and library stub at once. We
offer Axiomander as a data point that Iris's framing and ownership story is the
right substrate for \emph{automated}, composition-heavy verification, not only
for expert-driven interactive proof.

\section{Lessons learned}
\label{sec:lessons}

\begin{enumerate}
  \item \textbf{Framing is the fundamental abstraction.} Every composition
  property we needed reduced to it. Building it by hand over a flat store is where
  the WP blowup and the brittle automation came from; the substructural logic
  makes the cost vanish.
  \item \textbf{One boundary does many jobs.} Opaque specifications unify modular
  verification, proof reuse, and library stubbing -- and the incremental cache
  keys on the same contract hash. Designing them as one mechanism, rather than
  three, simplified the whole pipeline.
  \item \textbf{Incrementality follows contract boundaries.} Once callers depend
  only on contracts, ``what must I re-prove after this edit?'' has a syntactic
  answer (the changed contract hashes), and per-stage hashing refines it further.
  \item \textbf{Ownership is an expressivity question, not only efficiency.} Real
  Python mutates shared objects and raises through them. A flat store cannot
  express that, so a substructural logic is what makes such code verifiable at
  all -- the smaller proofs are a bonus.
  \item \textbf{Separation logic is easier to adopt than to reconstruct.} We
  spent more engineering faking the frame rule -- clobber operators, generated
  lemmas, \code{cbn} exclusions -- than we did inheriting it from Iris. If
  framing is load-bearing, reach for the substructural logic first.
\end{enumerate}

\paragraph{Status.} Snakelet and SnakeletWp are complete and free of
\code{Admitted} across the rule set of Appendix~\ref{app:snakelet}, including the
exception-aware extension and its loop rules; the staged generator, the
end-to-end Python-to-Iris pipeline, the contract-to-\code{iProp} compiler, and
the SMT-axiom slot are in place, with tens of Iris-specific tests passing
alongside the mature flat-store backend. Remaining work wires the exception
calculus into the Python lowering and extends the fragment to loop invariants over
shared disjoint memory, in-place data-structure mutation, and type-aware
reasoning over strings\slash floats\slash\code{isinstance}\slash\code{None}.

\paragraph{Open questions.} Two we would most value input on: how far an object's
footprint can be \emph{inferred} from \code{modifies} for an LLM-generated body
before the user must annotate \code{owns(...)}; and which Iris proof-mode
presentation of a failed stage's residual (hypothesis naming, spatial/persistent
split, masks elided) makes the best LLM prompt.

\section*{Artifact pointers}

This paper summarises material developed in the Axiomander repository: the Iris
migration plan and current state; the Iris backend prototype (the
\code{bump}/\code{owns} example and resource classification); the per-callee frame
lemmas in the flat-store backend (Section~\ref{sec:imp}); the staged proof
engineering notes; the incremental-verification cache design; and the companion
whitepaper and slides motivating vericoding.

\appendix

\section{The Snakelet calculus}
\label{app:snakelet}

Snakelet is an expression language in administrative-normal form: every operator
and call takes values or variables, and evaluation order is made explicit by
\code{let}-binding and a per-item evaluation-context grammar. A store is a finite
map from locations to values ($\sigma : \mathrm{loc} \rightharpoonup v$); a mutable
Python object field is a heap cell, exposed in the logic as $l \pto v$. The
immutable Python containers -- list, tuple, dict, set -- are \emph{structural}
values, not heap objects, mirroring Dafny's \code{seq} and F\textsuperscript{$\star$}'s
\code{list}.

Two coordinated developments share this design. The core language carries float
literals, dictionary get/set, and the concurrency primitives \code{FAA}/\code{fork};
the exception-aware extension shown here adds first-class exceptions and carries
the complete, machine-checked rule set. We present the latter because exceptions
change the \emph{shape of the judgement}. Following van
Collem/de~Vilhena/Krebbers (PLDI~2026), $\mathsf{raise}\;v$ on a value is a
\textbf{stuck terminal} expression -- not itself a value -- so an uncaught raise
sits in evaluation position as the program's exceptional result. A program
terminates in one of two ways, captured by the result type $r ::= \mathsf{val}\;v
\mid \mathsf{exn}\;s\;v$, and the WP postcondition ranges over $r$ rather than
over values. Figure~\ref{fig:grammar} gives the term language.

\begin{figure}[t]
\footnotesize
\[
\begin{array}{r@{\ }c@{\ }l}
v &::=& n \mid b \mid \mathit{fl} \mid s \mid l \mid () \mid \mathsf{exn}\;s\;v \\
  &\mid& [\,\overline{v}\,] \mid (\,\overline{v}\,) \mid \{\,\overline{v\!:\!v}\,\} \mid \{\,\overline{v}\,\} \\[2pt]
\mathit{op} &::=& {+} \mid {-} \mid {\times} \mid {/} \mid \mathbin{\%} \mid {=} \mid {\neq} \mid {<} \mid {\leq} \mid {>} \mid {\geq} \\
  &\mid& \mathsf{and} \mid \mathsf{or} \mid \mathsf{in} \mid \mathsf{len} \mid \mathsf{union} \mid \mathsf{inter} \\[2pt]
e &::=& v \mid x \mid \mathsf{let}\;x{=}e\;\mathsf{in}\;e \mid \mathit{op}(e,e) \mid g(\overline{e}) \\
  &\mid& \mathord{!}e \mid e \mathbin{:=} e \mid \mathsf{ref}\;e \mid \mathsf{if}\;e\;\mathsf{then}\;e\;\mathsf{else}\;e \\
  &\mid& \mathsf{raise}\;e \mid \mathsf{try}\;e\;\mathsf{catch}\;x{\Rightarrow}e \\
  &\mid& \mathsf{while}\;e\;e \mid \mathsf{for}\;x\;\mathsf{in}\;e\;\mathsf{do}\;e \\[2pt]
K &::=& \mathsf{let}\;x{=}\square\;\mathsf{in}\;e \mid \mathit{op}(\square,v) \mid \mathit{op}(e,\square) \mid \mathord{!}\square \mid \mathsf{ref}\;\square \\
  &\mid& \square \mathbin{:=} v \mid e \mathbin{:=} \square \mid \mathsf{if}\;\square\;\mathsf{then}\;e\;\mathsf{else}\;e \mid \mathsf{raise}\;\square \\
  &\mid& \mathsf{try}\;\square\;\mathsf{catch}\;x{\Rightarrow}e \mid \mathsf{for}\;x\;\mathsf{in}\;\square\;\mathsf{do}\;e \\[2pt]
r &::=& \mathsf{val}\;v \mid \mathsf{exn}\;s\;v
\end{array}
\]
\caption{Snakelet term language: values $v$ (integer, boolean, float, string,
location, unit, exception object, and the structural list/tuple/dict/set),
operators $\mathit{op}$, expressions $e$, single-item evaluation contexts $K$, and
results $r$. Every $K$ is \emph{neutral} except $\mathsf{try}\;\square\,\dots$;
$\mathit{op}(e,e)$ evaluates its right operand first. The core development
additionally provides $\mathsf{FAA}$, $\mathsf{fork}$, and dictionary get/set.}
\label{fig:grammar}
\end{figure}

\paragraph{Operational semantics.} Reduction is small-step and factors into
\emph{pure} steps $e \sstep e'$ (no heap, deterministic) and \emph{head} steps
$(e,\sigma) \rightarrow (e',\sigma';\vec{\mathit{ef}})$ (heap and calls), each
lifted under an evaluation context by \textsc{Ctx-P}/\textsc{Ctx-H}
(Figure~\ref{fig:opsem}). All context items are neutral except
$\mathsf{try}\;\square\;\mathsf{catch}\;x{\Rightarrow}h$; the unwinding rule fires
only for neutral $K$, so a raise propagates through \code{let}, operators, and
loads but is caught by an enclosing \code{try}. Calls dispatch through the
contract table $\entries$: $\mathsf{FunSpec}\;P\;Q$ (opaque -- stuck unless $P$
holds) or $\mathsf{FunDef}\;\vec{x}\;b$ (transparent -- substitute into the body).

\begin{figure*}[t]
\centering
\small
\textbf{Pure reduction}\quad $e \sstep e'$
\begin{mathpar}
\inferrule*[left=\textsc{Let}]{\,}{\mathsf{let}\;x{=}v\;\mathsf{in}\;e \sstep \subs{e}{x}{v}}
\and
\inferrule*[left=\textsc{Op}]{\,}{\mathit{op}(v_1,v_2) \sstep \binopd{v_1}{v_2}}
\and
\inferrule*[left=\textsc{IfT}]{\,}{\mathsf{if}\;\mathsf{true}\;\mathsf{then}\;e_1\;\mathsf{else}\;e_2 \sstep e_1}
\and
\inferrule*[left=\textsc{IfF}]{\,}{\mathsf{if}\;\mathsf{false}\;\mathsf{then}\;e_1\;\mathsf{else}\;e_2 \sstep e_2}
\and
\inferrule*[left=\textsc{TryV}]{\,}{\mathsf{try}\;v\;\mathsf{catch}\;x{\Rightarrow}h \sstep v}
\and
\inferrule*[left=\textsc{TryC}]{\,}{\mathsf{try}\;(\mathsf{raise}\;v)\;\mathsf{catch}\;x{\Rightarrow}h \sstep \subs{h}{x}{v}}
\and
\inferrule*[left=\textsc{Unwind}]{\mathrm{neutral}(K)}{K[\mathsf{raise}\;v] \sstep \mathsf{raise}\;v}
\and
\inferrule*[left=\textsc{While}]{\,}{\mathsf{while}\;e_1\;e_2 \sstep \mathsf{if}\;e_1\;\mathsf{then}\;(e_2 \mathbin{;} \mathsf{while}\;e_1\;e_2)\;\mathsf{else}\;()}
\and
\inferrule*[left=\textsc{ForNil}]{\,}{\mathsf{for}\;x\;\mathsf{in}\;[\,]\;\mathsf{do}\;b \sstep ()}
\and
\inferrule*[left=\textsc{ForCons}]{\,}{\mathsf{for}\;x\;\mathsf{in}\;(v\!::\!\mathit{vs})\;\mathsf{do}\;b \sstep \subs{b}{x}{v} \mathbin{;} \mathsf{for}\;x\;\mathsf{in}\;\mathit{vs}\;\mathsf{do}\;b}
\end{mathpar}
\smallskip
\textbf{Head reduction}\quad $(e,\sigma) \rightarrow (e',\sigma';\vec{\mathit{ef}})$ \qquad and \quad \textbf{context decomposition}
\begin{mathpar}
\inferrule*[left=\textsc{Load}]{\sigma(l)=v}{(\mathord{!}l,\sigma) \rightarrow (v,\sigma)}
\and
\inferrule*[left=\textsc{Store}]{l \in \mathrm{dom}\,\sigma}{(l \mathbin{:=} v,\sigma) \rightarrow ((),\,\sigma[l{\mapsto}v])}
\and
\inferrule*[left=\textsc{Alloc}]{\sigma(l)=\bot}{(\mathsf{ref}\;v,\sigma) \rightarrow (l,\,\sigma[l{\mapsto}v])}
\and
\inferrule*[left=\textsc{CallSpec}]{\entries(g)=\mathsf{FunSpec}\;P\;Q \\ P\,\vec{v} \\ Q\,\vec{v}\,v}{(g(\vec{v}),\sigma) \rightarrow (v,\sigma)}
\and
\inferrule*[left=\textsc{CallDef}]{\entries(g)=\mathsf{FunDef}\;\vec{x}\;b \\ |\vec{v}|=|\vec{x}|}{(g(\vec{v}),\sigma) \rightarrow (\subs{b}{\vec{x}}{\vec{v}},\sigma)}
\and
\inferrule*[left=\textsc{Ctx-P}]{e \sstep e'}{(K[e],\sigma) \rightarrow (K[e'],\sigma)}
\and
\inferrule*[left=\textsc{Ctx-H}]{(e,\sigma) \rightarrow (e',\sigma';\vec{\mathit{ef}})}{(K[e],\sigma) \rightarrow (K[e'],\sigma';\vec{\mathit{ef}})}
\end{mathpar}
\caption{Snakelet operational semantics. Pure steps are heap-independent and
deterministic; head steps touch the heap or the contract table $\entries$.
\textsc{CallSpec} fires only inside the callee's precondition $P$ (calling outside
$P$ is stuck), making the precondition a proof obligation on the caller.
$\vec{\mathit{ef}}$ collects forked threads (empty except for \code{fork} in the
core development).}
\label{fig:opsem}
\end{figure*}

\paragraph{Weakest preconditions.} SnakeletWp's postcondition is $\Phi :
\code{Result} \to \mathit{iProp}$. It is the guarded fixpoint of: if $e$ is
terminal then $\Phi\,r$ holds under a fancy update; otherwise $e$ is reducible and,
after one step, the state interpretation (the Iris \code{gen\_heap} authoritative
view) and the WP of the reduct are re-established. Ranging $\Phi$ over
\code{Result} is the one structural change exceptions force. Composition goes
through \textsc{Bind}, which threads a \code{Result}-typed continuation
$\mathrm{bindp}\,K\,\Phi$ with $\mathrm{bindp}\,K\,\Phi\,(\rval{v}) =
\wpe{K[v]}{\Phi}$ and $\mathrm{bindp}\,K\,\Phi\,(\rexn{s}{v}) = \Phi(\rexn{s}{v})$
-- on a value it continues, on an exception it short-circuits to the exceptional
postcondition. The heap rules are the usual Iris points-to rules; their
exceptional arm describes the heap \emph{as it stood at the raise}. On top of
\textsc{ForCons} we derive a structural fold rule carrying a user invariant $P$
over the remaining list suffix:
\[
\begin{array}{l}
P\,M \;\sepc\; \big(\,\Box\,\forall v\,\mathit{vs}.\ P\,(v\!::\!\mathit{vs}) \wand
  \wpe{\subs{b}{x}{v}}{r.\,\theta(r)}\,\big) \\[2pt]
\quad {}\sepc\; (P\,[\,] \wand \Phi(\rval{()})) \;\;\vdash\;\;
  \wpe{\mathsf{for}\;x\;\mathsf{in}\;M\;\mathsf{do}\;b}{\Phi},
\end{array}
\]
where $\theta(\rval{\_}) = P\,\mathit{vs}$ and $\theta(\rexn{s}{v}) =
\Phi(\rexn{s}{v})$. Each rule has a matching stage tactic
(Section~\ref{sec:staged}): \code{pure\_step} fires
\textsc{Let}\slash\textsc{Op}\slash\textsc{IfT}\slash\textsc{IfF}\slash\textsc{TryV}\slash\textsc{TryC},
\code{heap\_load}\slash\code{store}\slash\code{alloc} the heap rules,
\code{call\_opaque}\slash\code{call\_transparent} the call rules, \code{raise\_step}
\textsc{Raise}\slash\textsc{Unwind}, and \code{loop\_unfold} the loop rules.

\begin{figure*}[t]
\centering
\small
\begin{mathpar}
\inferrule*[left=\textsc{Val}]{\,}{\Phi(\rval{v}) \vdash \wpe{v}{\Phi}}
\and
\inferrule*[left=\textsc{Raise}]{\,}{\Phi(\rexn{s}{v}) \vdash \wpe{\mathsf{raise}\;(\rexn{s}{v})}{\Phi}}
\and
\inferrule*[left=\textsc{Wand}]{\,}{\wpe{e}{\Phi} \sepc (\forall r.\ \Phi\,r \wand \Psi\,r) \vdash \wpe{e}{\Psi}}
\and
\inferrule*[left=\textsc{Let}]{\,}{\later \wpe{\subs{e}{x}{v}}{\Phi} \vdash \wpe{\mathsf{let}\;x{=}v\;\mathsf{in}\;e}{\Phi}}
\and
\inferrule*[left=\textsc{LetRaise}]{\,}{\Phi(\rexn{s}{v}) \vdash \wpe{\mathsf{let}\;x{=}\mathsf{raise}\,(\rexn{s}{v})\;\mathsf{in}\;e}{\Phi}}
\and
\inferrule*[left=\textsc{Op}]{\,}{\later \wpe{\binopd{v_1}{v_2}}{\Phi} \vdash \wpe{\mathit{op}(v_1,v_2)}{\Phi}}
\and
\inferrule*[left=\textsc{IfT}]{\,}{\later \wpe{e_1}{\Phi} \vdash \wpe{\mathsf{if}\;\mathsf{true}\;\mathsf{then}\;e_1\;\mathsf{else}\;e_2}{\Phi}}
\and
\inferrule*[left=\textsc{IfF}]{\,}{\later \wpe{e_2}{\Phi} \vdash \wpe{\mathsf{if}\;\mathsf{false}\;\mathsf{then}\;e_1\;\mathsf{else}\;e_2}{\Phi}}
\and
\inferrule*[left=\textsc{TryV}]{\,}{\later \wpe{v}{\Phi} \vdash \wpe{\mathsf{try}\;v\;\mathsf{catch}\;x{\Rightarrow}h}{\Phi}}
\and
\inferrule*[left=\textsc{TryC}]{\,}{\later \wpe{\subs{h}{x}{v}}{\Phi} \vdash \wpe{\mathsf{try}\;(\mathsf{raise}\;v)\;\mathsf{catch}\;x{\Rightarrow}h}{\Phi}}
\and
\inferrule*[left=\textsc{Bind}]{\mathrm{neutral}(K)}{\wpe{e}{\mathrm{bindp}\,K\,\Phi} \vdash \wpe{K[e]}{\Phi}}
\and
\inferrule*[left=\textsc{Load}]{\,}{l \pto v \;\sepc\; \later (l \pto v \wand \Phi(\rval{v})) \vdash \wpe{\mathord{!}l}{\Phi}}
\and
\inferrule*[left=\textsc{Store}]{\,}{l \pto w \;\sepc\; \later (l \pto v \wand \Phi(\rval{()})) \vdash \wpe{l \mathbin{:=} v}{\Phi}}
\and
\inferrule*[left=\textsc{Alloc}]{\,}{\later (\forall l.\ l \pto v \wand \Phi(\rval{l})) \vdash \wpe{\mathsf{ref}\;v}{\Phi}}
\and
\inferrule*[left=\textsc{ForNil}]{\,}{\later \Phi(\rval{()}) \vdash \wpe{\mathsf{for}\;x\;\mathsf{in}\;[\,]\;\mathsf{do}\;b}{\Phi}}
\and
\inferrule*[left=\textsc{CallOpaque}]{\entries(g)=\mathsf{FunSpec}\;P\;Q \\ P\,\vec{v}}{\later (\forall v.\ \pure{Q\,\vec{v}\,v} \wand \Phi(\rval{v})) \vdash \wpe{g(\vec{v})}{\Phi}}
\and
\inferrule*[left=\textsc{CallDef}]{\entries(g)=\mathsf{FunDef}\;\vec{x}\;b \\ |\vec{v}|=|\vec{x}|}{\later \wpe{\subs{b}{\vec{x}}{\vec{v}}}{\Phi} \vdash \wpe{g(\vec{v})}{\Phi}}
\end{mathpar}
\[
\begin{array}{l}
\textsc{While}\quad \later \wpe{\mathsf{if}\;e_1\;\mathsf{then}\;(e_2 \mathbin{;} \mathsf{while}\;e_1\;e_2)\;\mathsf{else}\;()}{\Phi} \\
\qquad\quad {}\vdash\; \wpe{\mathsf{while}\;e_1\;e_2}{\Phi} \\[4pt]
\textsc{ForCons}\quad \later \wpe{\subs{b}{x}{v} \mathbin{;} \mathsf{for}\;x\;\mathsf{in}\;\mathit{vs}\;\mathsf{do}\;b}{\Phi} \\
\qquad\quad {}\vdash\; \wpe{\mathsf{for}\;x\;\mathsf{in}\;(v\!::\!\mathit{vs})\;\mathsf{do}\;b}{\Phi}
\end{array}
\]
\caption{SnakeletWp weakest-precondition rules. The postcondition $\Phi :
\code{Result} \to \mathit{iProp}$ ranges over $r ::= \rval{v} \mid \rexn{s}{v}$;
$\later$ is the Iris later modality, $\sepc$ the separating conjunction, $\wand$
the magic wand, and $\pure{\cdot}$ the persistent embedding of a pure proposition.
\textsc{Bind} uses the \code{Result}-transformer $\mathrm{bindp}$ (text); the
wider \textsc{While}/\textsc{ForCons} rules are shown beneath the grid. Every
rule is machine-checked.}
\label{fig:wp}
\end{figure*}

\end{document}
